How to publish in Biological Sciences: a guide for the uninitiated

### 2023-06-20 00-author.Rmd
Add files via upload
# new chapter on pub bias

Measey, J. (`r format(Sys.time(), '%Y')`) _How to publish in Biological Sciences: a guide for the uninitiated_. Available from: https://howtopublishscience.org/ Accessed on: `r format(Sys.time(), '%d %B %Y')`.

## Creative Commons License {-}
"**How to publish in Biological Sciences: A guide for the uninitiated**" is copyright to John Measey and distributed under a Creative Commons BY-NC-ND 4.0 International License. You may freely share this content for non-commercial purposes, with a source credit to www.john.measey.com

>>>>

Measey, J. (`r format(Sys.time(), '%Y')`) _How to publish in Biological Sciences: a guide for the uninitiated_. Available from: https://howtopublishscience.org/ Updated on: `r format(Sys.time(), '%d %B %Y')`.

## Creative Commons License {-}
"**How to publish in Biological Sciences: A guide for the uninitiated**" is copyright to John Measey and distributed under a [Creative Commons BY-NC-ND 4.0 International License](https://creativecommons.org/licenses/by-nc-nd/4.0/). You may freely share this content for non-commercial purposes, with a source credit to [howtopublishscience.org](https://howtopublishscience.org/)

```{r fig.align='centre', out.width = '66%', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='https://creativecommons.org/licenses/by-nc-nd/4.0/'}
knitr::include_graphics('figures/BY-NC-ND.png', dpi = NA)
```

### 2023-06-20 1.09-preprints.Rmd
# new chapter on APCs
Add files via upload
# new chapter on pub bias

According to Brown [-@brown2010introduction], the idea of overlaying has been with us for some time, and exists as websites that offer a series of links to other papers. In this way, a review article could be considered an 'overlay paper', the contents of Web of Science as an 'overlay database'. But, for me at least, this is not where the real potential lies. Instead, imagine the overlay journal as a way in which academics entirely remove the need for publishers. The need for this is increasingly evident as we become more familiar with the ways in which we rely on traditional publishing models to pervade our scientific project with confirmation bias. Overlay journals no longer require a publisher to store the publication. This is done at the preprint server. The reviews are housed at the same *arXiv* site (or would be in an ideal and transparent version) [@rittman2020fast], as is the manuscript in its final form after being accepted by the overlay journal editor. The authors themselves are responsible for the final layout. The Overlay Journal co-ordinates the reviews and conducts the editorial work, and then simply acts as a pointer to the finished product: no papers, no publishers, no editorial management software, no costs and all papers are [Diamond OA](#diamond)!

@@

One of the original electronic journals, [*Journal of Medical Internet Research* (*JMIR*)](https://www.jmir.org/), announced in 2019 that it will launch an overlay journal covering biology, a so called 'superjournal', *JMIRx* [@eysenbach2019celebrating]. However, JMIR appears to be exploiting the name 'overlay journal' without adhering to any of the qualities - they are essentially more [Gold OA](#gold) journals with ['no impact factor'](#noif) aspirations that intend to use the [Article Processing Charge](#APC) to finance their operation. Editors may use preprints to approach authors of papers that they want to publish ('editorial prospecting'), and then approaching reviewers, and also by authors pitching their preprints to the editors. Today [*JMIRx|Bio*](https://bio.jmirx.org/) accepts any preprint published on [*bioRxiv*](#bioRxiv). Although *JMIRx|Bio*, and sister journal [*JMRIx|Psy*](https://psy.jmirx.org/), were launched in 2020, I cannot find any articles submitted (by mid-2022). The sister journal [*JMIRx|Med*](https://med.jmirx.org/), launched at the same time and in the same area as other JMIR journals, and has a rapidly expanding publication base.  But if you want to publish in JMIR you can expect to pay an [APC](#APC), and so they are not overlay journals as they archive their own content. But JMIR is disrupting the publishing scene and offers a lot of benefits towards open science.

@@

Another initiative from _Peer Community In_ is the possibility of submitting to their [Registered Reports](https://rr.peercommunityin.org/), which goes much further towards removing the confirmation bias. The Registered Report (RR) is in effect the registration of a proposal (i.e. preregistration) with review. If the RR is approved by reviewers then the study is, in principle, given the green light for publication whether or not the hypothesis posed is accepted or rejected. I say “in principle” because those same reviewers are shown the manuscript again once the results are in. They need to check that the methods proposed were followed, the analyses were conducted in the same way they were proposed, and that the conclusions are justified by the results. Peer Community In are offering to organise the two sets of peer review. In addition, there are a bunch of journals that have already signed up to accept RRs that are signed off after completion (notable among these is PeerJ). To me, this represents an important step in the right direction toward [transparency](#transparency2), and the elimination of confirmation bias. What would be great to see is the number of conventional journals sign up with RRs based on the quality of the study design and execution, and the concomitant abandonment of [Impact Factor](#impactfactor) as a driving force in publishing.

>>>>

According to Brown [-@brown2010introduction], the idea of overlaying has been with us for some time, and exists as websites that offer a series of links to other papers. In this way, a review article could be considered an 'overlay paper', the contents of Web of Science as an 'overlay database'. But, for me at least, this is not where the real potential lies. Instead, imagine the overlay journal as a way in which academics entirely remove the need for publishers. The need for this is increasingly evident as we become more familiar with the ways in which we rely on traditional publishing models to pervade our scientific project with [confirmation bias](#confirmation). Overlay journals no longer require a publisher to store the publication. This is done at the preprint server. The reviews are housed at the same *arXiv* site (or would be in an ideal and transparent version) [@rittman2020fast], as is the manuscript in its final form after being accepted by the overlay journal editor. The authors themselves are responsible for the final layout. The Overlay Journal co-ordinates the reviews and conducts the editorial work, and then simply acts as a pointer to the finished product: no papers, no publishers, no editorial management software, no costs and all papers are [Diamond OA](#diamond)!

@@

One of the original electronic journals, [*Journal of Medical Internet Research* (*JMIR*)](https://www.jmir.org/), announced in 2019 that it will launch an overlay journal covering biology, a so called 'superjournal', *JMIRx* [@eysenbach2019celebrating]. However, JMIR appears to be exploiting the name 'overlay journal' without adhering to any of the qualities - they are essentially more [Gold OA](#gold) journals with ['no impact factor'](#noif) aspirations that intend to use the [Article Processing Charge](#APC) to finance their operation. Editors may use preprints to approach authors of papers that they want to publish ('editorial prospecting'), and then approaching reviewers, and also by authors pitching their preprints to the editors. Today [*JMIRx|Bio*](https://bio.jmirx.org/) accepts any preprint published on [*bioRxiv*](#bioRxiv). Although *JMIRx|Bio*, and sister journal [*JMRIx|Psy*](https://psy.jmirx.org/), were launched in 2020, I cannot find any articles submitted (by mid-2022). The sister journal [*JMIRx|Med*](https://med.jmirx.org/), launched at the same time and in the same area as other JMIR journals, and has a rapidly expanding publication base.  But if you want to publish in JMIR you can expect to pay an [APC](#APC), and so they are not overlay journals as they archive their own content. But JMIR is disrupting the publishing scene and offers a lot of benefits towards [open science](#openscience).

@@

Another initiative from _Peer Community In_ is the possibility of submitting to their [Registered Reports](https://rr.peercommunityin.org/), which goes much further towards removing the [confirmation bias](#confirmation). The Registered Report (RR) is in effect the registration of a proposal (i.e. preregistration) with review. If the RR is approved by reviewers then the study is, in principle, given the green light for publication whether or not the hypothesis posed is accepted or rejected. I say “in principle” because those same reviewers are shown the manuscript again once the results are in. They need to check that the methods proposed were followed, the analyses were conducted in the same way they were proposed, and that the conclusions are justified by the results. Peer Community In are offering to organise the two sets of peer review. In addition, there are a bunch of journals that have already signed up to accept RRs that are signed off after completion (notable among these is PeerJ). To me, this represents an important step in the right direction toward [transparency](#transparency2), and the elimination of [confirmation bias](#confirmation). What would be great to see is the number of conventional journals sign up with RRs based on the quality of the study design and execution, and the concomitant abandonment of [Impact Factor](#impactfactor) as a driving force in publishing.

### 2023-06-20 2.04-whichjournal.Rmd
Add files via upload
# some corrections & additions
Add files via upload
# new chapter on pub bias

### Knowing the journal from the inside
Your mentor or co-authors could be an Editor or Associate Editor (past or present) of one of your target journals. Or there may be someone in your department or institute that you could consult. I am not suggesting that you use their influence to help you get published, this conflict of interest strictly prohibited by most journals. Instead, these people can help you decide whether or not your submission will be welcomed or [desk rejected](#desk).

@@

There's no problem to publish in a journal that has no interest in Impact Factor. Indeed, it can be seen as 'the right thing to do', with respect to confirmation bias. This is the way that all science was published prior to the dominance of [citation metrics](#citations). If you are submitting to a 'no impact factor journal' then you should be aware that this is what their speciality is: an emphasis on the technical soundness, rather than whether or not a significant result was found. You need to be aware as this is how inclusion of such publications in your CV will be regarded by peers. The list of 'no impact factor journals' is growing, and they tend to be online only, [Gold OA](#gold) and carry substantial [article processing charges](#gold).

>>>>

### Knowing the journal from the inside
Your mentor or co-authors could be an Editor or Associate Editor (past or present) of one of your target journals. Or there may be someone in your department or institute that you could consult. I am not suggesting that you use their influence to help you get published, this [conflict of interest](#conflicts) strictly prohibited by most journals. Instead, these people can help you decide whether or not your submission will be welcomed or [desk rejected](#desk).

@@

There's no problem to publish in a journal that has no interest in Impact Factor. Indeed, it can be seen as 'the right thing to do', with respect to [confirmation bias](#confirmation). This is the way that all science was published prior to the dominance of [citation metrics](#citations). If you are submitting to a 'no impact factor journal' then you should be aware that this is what their speciality is: an emphasis on the technical soundness, rather than whether or not a significant result was found. You need to be aware as this is how inclusion of such publications in your CV will be regarded by peers. The list of 'no impact factor journals' is growing, and they tend to be online only, [Gold OA](#gold) and carry substantial [article processing charges](#gold).

### 2023-06-20 4.06-publishmore.Rmd
Add files via upload
# new chapter on pub bias

(ref:authors-writing-more) **The growth in the number of papers published in Life Sciences over time.** The number of papers published (blue line) compared to a standard growth rate (black line). The data come from www.scopus.com.

@@

But it is possible that these figures could be explained by the fact that the culture in publishing in biological sciences has changed a great deal since then. One hundred years ago, it was very unlikely that any postgraduate students would publish articles in peer reviewed journals. Moreover, it was also acceptable for advisers to take the thesis work of their students and write it up in monographs. This has certainly changed with the ranks of authors now being swelled considerably so that many more authors are likely to be included on only a single publication in which they participated [see @measey2011past]. I interpret this as the biological sciences becoming more democratic, with more of the people that contribute to the work receiving the [credit](#authors).

@@

@@

### Natural selection of bad science {#badscience}
In -@smaldino2016natural, Smaldino and McElreath proposed that ever increasing numbers of publications not only leads to bad science, but that we should expect to see this in an academic environment where publishing is considered as a currency. They argued that the most productive laboratories will be rewarded with more grant funding, larger numbers of students, and that these students will learn about the methods and benefits of prolific publication. When these ‘offspring’ of the prolific lab look for jobs, they are more likely to be successful as they have more publications themselves. An academic environment that rewards increasing numbers of publications eventually selects towards methodologies that produce the greatest number of publishable results. To show that this leads to a culture of ‘bad science’, Smaldino and McElreath [-@smaldino2016natural] conducted an analysis in trends over time of statistical power in behavioural science publications. Over time, better science should be shown by researchers increasing their statistical power as this will provide studies with lower error rates. However, increasing the statistical power of experiments takes more time and resources, resulting in fewer publications. Their results, from review papers in social and behavioural sciences, suggested that between 1960 and 2011 there had been no trend toward increasing statistical power. Biological systems, whether they be academics in a department or grass growing in experimental pots, will respond to the rewards generated in that system. When grant funding bodies and academic institutions reward publishing as a behaviour, it is inevitable that the behaviour of researchers inside that system will respond by increasing their publication output. Moreover, if those institutions maintain increasing numbers of researchers in temporary positions, those individuals are further incentivised to become more productive to justify their continued contracts, or the possibility of obtaining a (more permanent) position elsewhere. Eventually, this negative feedback, or gamification of publishing metrics, produces a dystopian and dysfunctional academic reality [@chapman2019games; @helmer2020what].

An example of this kind of confirmation bias driven publishing effect toward bad science can be found in the literature of fluctuating asymmetry, and in particular those studies on human faces [@dongen2011associations]. Back in the 1990s, there was a flurry of high profile articles purporting preference for symmetry (and against asymmetry) in human faces. The studies were (relatively) cheap and fast to conduct as the researchers had access to hundreds of students right on their doorsteps. The studies not only hit the top journals, but were very popular in the mainstream media as scientists were apparently able to predict which faces were the most attractive.

Stefan van Dongen [-@dongen2011associations] hypothesised that if publication bias was leading to bad science in studies of fluctuating asymmetry in human faces, there would be a negative association between effect size and sample size. However, effect sizes can also be expected to be smaller in larger studies as these may come with less accurate measurements [see @jennions2002publication]. This negative association should not change depending on the stated focus of the study; i.e. if the result was the primary or secondary outcome. However, van Dongen [-@dongen2011associations] found that in studies where the relationship between fluctuating asymmetry and attractiveness was a secondary outcome (not the primary focus of the study), the overall effect size did not differ from zero and no association between effect size and sample size was found. This was in contrast to the studies where the fluctuating asymmetry-attractiveness was a primary focus, suggesting that there was important publication bias in this area of research.

Where others have looked, publication bias has been found and is particularly associated with a decreasing effect size that correlates with journal [Impact Factor](#impactfactor): i.e. once the large effect is published in a big journal, the natural selection of bad science results in publication bias, and diminishing effect sizes that ripple through lower impact factor journals [@munafo2007association; @brembs2013deep; @smaldino2016natural], while negative results disappear almost entirely [@fanelli2012negative]. One can presume that negative results exist, but their authors either do not bother to write up the results, or that journals won't publish them.

>>>>

(ref:authors-writing-more) **The growth in the number of papers published in Life Sciences over time.** The number of papers published (blue line) compared to a standard growth rate (black line) shows an exponential and not a linear increase in Life Sciences. The data come from www.scopus.com.

@@

But it is possible that these figures could be explained by the fact that the culture in publishing in biological sciences has changed a great deal. One hundred years ago, it was very unlikely that any postgraduate students would publish articles in peer reviewed journals. Moreover, it was also acceptable for advisors to take the thesis work of their students and write it up in monographs. This has certainly changed with the ranks of authors now being swelled considerably so that many more authors are likely to be included on only a single publication in which they participated [see @measey2011past]. I interpret this as the biological sciences becoming more democratic, with more of the people that contribute to the work receiving the [credit](#authors).

@@

The data (from Scopus) in Figure \@ref(fig:authors-writing-more) suggest that in the Life Sciences, the doubling time is as little as 10 years: twice as much in 2020 as there was in 2010. The previous doubling period was 20 years, between 1990 and 2010. This exponential rate of increasing literature appears higher than in other subjects.

@@

### 2023-06-20 4.07-retract.Rmd
Add files via upload
# new chapter on pub bias

A retraction of a paper is when your paper is effectively 'unpublished'. This happens at the discretion of the editor (and often the entire editorial board), and is a very serious issue. Retractions are rare. Reasons for retractions vary. It could be that a piece of equipment was later found to have malfunctioned or was calibrated incorrectly [@anon2018retraction]. A cell line was misidentified. Or they can be through no fault of the authors. For example, Toro et al. [-@toro2019retraction] had their manuscript rejected by _Journal of Biosciences_, but due to an administrative error, the article was printed in an issue, and later retracted. However, the top reason for retraction is now misconduct [@fang2012misconduct; @brainard2018what], and this is hardly surprising given the crazy incentives that many scientists received to publish in journals with top impact factors. Another important factor with retractions is that they appear to be more common in journals with higher [impact factors](#impactfactor) [@brembs2013deep], and this should not surprise us as these journals are prone to publishing studies with [confirmation bias](http://www.howtowriteaphd.org/TypeI.html) [@forstmeier2017detecting; @measey2021how].

>>>>

A retraction of a paper is when your paper is effectively 'unpublished'. This happens at the discretion of the editor (and often the entire editorial board), and is a very serious issue. Retractions are rare. Reasons for retractions vary. It could be that a piece of equipment was later found to have malfunctioned or was calibrated incorrectly [@anon2018retraction]. A cell line was misidentified. Or they can be through no fault of the authors. For example, Toro et al. [-@toro2019retraction] had their manuscript rejected by _Journal of Biosciences_, but due to an administrative error, the article was printed in an issue, and later retracted. However, the top reason for retraction is now misconduct [@fang2012misconduct; @brainard2018what], and this is hardly surprising given the crazy incentives that many scientists received to publish in journals with top impact factors. Another important factor with retractions is that they appear to be more common in journals with higher [impact factors](#impactfactor) [@brembs2013deep], and this should not surprise us as these journals are prone to publishing studies with [confirmation bias](#confirmation) [@forstmeier2017detecting; @measey2021how].
